---
title: "Data Manipulation"
author: "Linh Tran"
date: "9/23/2020"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
```

## Loading data and cleaning column names

```{r loading_data}
options (tibble.print_min = 3)

litters_df = read_csv("data/FAS_litters.csv", 
                      col_types = "ccddiiii")
litters_df = clean_names(litters_df)

pups_df = read_csv("data/FAS_pups.csv",
                   col_types = "ciiiii")
pups_df = clean_names(pups_df)
```

## `select` - operate on columns

Choose some columns and not others.

```{r}
select(litters_df, group, gd0_weight)
```

Select from gd0_weight to gd_of_birth
```{r}
select(litters_df, group, gd0_weight: gd_of_birth)
```

Specify what we want to lose (litter number)
```{r}
select(litters_df, -litter_number)
```

Renaming columns
```{r}
select(litters_df, gd_weight_0 = gd0_weight, gd_weight_18 = gd18_weight)
```

Can also use rename()
```{r}
rename(litters_df, gd_weight_0 = gd0_weight, gd_weight_18 = gd18_weight)
```

Select helpers

```{r}
select(litters_df, starts_with("gd"))
```

```{r}
select(litters_df, litter_number, everything())
```

Relocate litter number columne to the beginning
```{r}
relocate(litters_df, litter_number)
```

## `filter` - operate on rows

```{r}
filter (litters_df, gd0_weight < 22)
```

```{r}
filter(litters_df, gd_of_birth == 20)
```

```{r}
filter(litters_df, pups_dead_birth == 0)
```

```{r}
filter(litters_df, gd0_weight <=20)
```

Negate things we are interested in

```{r}
filter(litters_df, !(gd_of_birth == 20))
filter(litters_df, gd_of_birth !=20)
```

```{r}
filter(litters_df, gd0_weight >=22, gd_of_birth == 20)
filter(litters_df, !(gd0_weight >=22), !(gd_of_birth == 20))
```

```{r}
filter(litters_df, group == "Con7")
filter(litters_df, group %in% c("Con7", "Mod8"))
filter(litters_df, group %in% c("Con7", "Mod8"), gd_of_birth == 20)
```

## `mutate`- allows you to modify existing variables or create new variables

```{r}
mutate(litters_df, wt_gain = gd18_weight - gd0_weight)
```

adding multiple variables at the same time

```{r}
mutate (
  litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```

## `arrange` - putting things in order

```{r}
arrange(litters_df, pups_born_alive)
arrange(litters_df, gd0_weight, gd18_weight)
```

## Pipe operator to tie together sequence of actions `%>%`

This is one option - would NOT recommend
```{r}
litters_data_raw = read_csv("data/FAS_litters.csv")
litters_clean_name = janitor::clean_names(litters_data_raw)
litters_data_selected = select(litters_clean_name, -pups_survive)
litters_mutated = mutate (litters_data_selected, wt_gain = gd18_weight - gd0_weight)
litters_without_missing_data = drop_na(litters_mutated, gd0_weight)
```

USE THE PIPE OPERATOR INSTEAD!!!

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv") %>% 
  janitor::clean_names() %>% 
  select(-pups_survive) %>% 
  mutate(wt_gain = gd18_weight - gd0_weight) %>% 
  drop_na(gd0_weight)
```

