---
title: "Tidy Data"
author: "Linh Tran"
date: "9/24/2020"
output: html_document
---

```{r setup}
library(tidyverse)
library(dplyr)
library(haven)
```

Load the PULSE data

```{r}
pulse_data =
  haven::read_sas("data/public_pulse_data.sas7bdat") %>% 
  janitor::clean_names()
```

## `pivot_longer` - wide format to long format

```{r}
pulse_data_tidy = 
  pulse_data %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  )
pulse_data_tidy
```

Changing `bl` to `00m` for consistency across visits, converting `visit` to a numeric variable using `mutate`

```{r}
pulse_data =
  haven::read_sas("data/public_pulse_data.sas7bdat") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  ) %>% 
  relocate(id, visit) %>% 
  mutate(visit = recode (visit, "bl" = "00m"))
pulse_data
print(pulse_data, n = 12)
```

## Learning Assessment - how to use `pivot_longer`

```{r}
litters_long = read_csv("data/FAS_litters.csv", col_types = "ccddiiii") %>% 
  janitor::clean_names() %>% 
  select (litter_number, ends_with ("weight")) %>% 
  pivot_longer(
    gd0_weight : gd18_weight,
    names_to = "gd",
    values_to = "weight") %>% 
  mutate(gd = recode(gd, "gd0_weight" = 0, "gd18_weight" = 18))
litters_long
```

## `pivot_wider` 

Let's take a short digression into untidying data. The code below creates a tidy dataset that could result from an analysis. 

```{r}
analysis_result = 
  tibble(
    group = c("treatment", "treatment", "placebo", "placebo"),
    time = c("pre", "post", "pre", "post"),
    mean = c(4, 8, 2, 1.5)
  )

print(analysis_result)

analysis_result %>% 
  pivot_wider(
    names_from = "time",
    values_from = "mean"
  )
```

## Binding rows

Using the LotR data

First step: import each table and add a variable to each dataframe indicating the movie. 

```{r}
fellowship_ring = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "B3:D6") %>% 
  mutate(movie = "fellowship_ring")
```

```{r}
two_towers = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "F3:H6") %>% 
  mutate(movie = "two_towers")
```

```{r}
return_king = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "J3:L6") %>% 
  mutate(movie = "return_king")
```

Stack the tables together using `bind_rows` and tidy the result

```{r binding_rows}
lotr_tidy = 
  bind_rows(fellowship_ring, two_towers, return_king) %>% 
  janitor::clean_names() %>% 
  relocate(movie) %>% 
  pivot_longer(
    female:male,
    names_to = "gender",
    values_to = "words"
  )
print(lotr_tidy)
```

## Joining datasets - combining two tables (Combining 3 or more is done by step-by-step using the same ideas)

Import and clean the FAS datasets.

```{r}
pups_df =
  read_csv("data/FAS_pups.csv", col_types = "ciiiii") %>% 
  janitor::clean_names() %>% 
  mutate(sex = recode(sex, `1` = "male", `2` = "female"))

litters_df =
  read_csv("data/FAS_litters.csv", col_types = "ccddiiii") %>% 
  janitor::clean_names() %>%  
  relocate(litter_number) %>% 
  separate(group, into = c("dose", "day_of_tx"), sep = 3) %>% 
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    dose = str_to_lower(dose)
  )

fas_data = left_join(pups_df, litters_df, by = "litter_number")
fas_data
```

Next up, time to join them!

```{r}
fas_df = 
  left_join(pups_df, litters_df, by = "litter_number") %>% 
  arrange(litter_number) %>% 
  relocate(litter_number, dose, day_of_tx)
```

## Learning Assessment - Joining datasets

```{r}
surv_os = read_csv("data/surv_os.csv", col_types = "cc") %>% 
  rename("UNI" = "What is your UNI?",  "operating_system" = "What operating system do you use?") 

surv_program_git = read_csv("data/surv_program_git.csv", col_types = "ccc") %>% 
  rename("UNI" = "What is your UNI?",
         "degree" = "What is your degree program?", 
         "experience_with_git" = "Which most accurately describes your experience with Git?")

surv_data_left = left_join (surv_os, surv_program_git, by = "UNI")
surv_data_right = right_join(surv_os, surv_program_git)
surv_data_inner = inner_join(surv_os, surv_program_git)
```

Solution

```{r}
surv_os = read_csv("data/surv_os.csv") %>% 
  janitor::clean_names() %>% 
  rename(id = what_is_your_uni, os = what_operating_system_do_you_use)

surv_pr_git = read_csv("data/surv_program_git.csv") %>% 
  janitor::clean_names() %>% 
  rename(
    id = what_is_your_uni, 
    prog = what_is_your_degree_program,
    git_exp = which_most_accurately_describes_your_experience_with_git)

left_join(surv_os, surv_pr_git)
inner_join(surv_os, surv_pr_git)
anti_join(surv_os, surv_pr_git)
anti_join(surv_pr_git, surv_os)

```

