---
title: "Tidy Data"
author: "Linh Tran"
date: "9/24/2020"
output: html_document
---

```{r setup}
library(tidyverse)
library(dplyr)
library(haven)
```

Load the PULSE data

```{r}
pulse_data =
  haven::read_sas("data/public_pulse_data.sas7bdat") %>% 
  janitor::clean_names()
```

## `pivot_longer` - wide format to long format

```{r}
pulse_data_tidy = 
  pulse_data %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  )
pulse_data_tidy
```

Changing `bl` to `00m` for consistency across visits, converting `visit` to a numeric variable using `mutate`

```{r}
pulse_data =
  haven::read_sas("data/public_pulse_data.sas7bdat") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  ) %>% 
  relocate(id, visit) %>% 
  mutate(visit = recode (visit, "bl" = "00m"))
pulse_data
print(pulse_data, n = 12)
```

## Learning Assessment - how to use `pivot_longer`

```{r}
litters_long = read_csv("data/FAS_litters.csv", col_types = "ccddiiii") %>% 
  janitor::clean_names() %>% 
  select (litter_number, ends_with ("weight")) %>% 
  pivot_longer(
    gd0_weight : gd18_weight,
    names_to = "gd",
    values_to = "weight") %>% 
  mutate(gd = recode(gd, "gd0_weight" = 0, "gd18_weight" = 18))
litters_long
```

## `pivot_wider` 

Let's take a short digression into untidying data. The code below creates a tidy dataset that could result from an analysis. 

```{r}
analysis_result = 
  tibble(
    group = c("treatment", "treatment", "placebo", "placebo"),
    time = c("pre", "post", "pre", "post"),
    mean = c(4, 8, 2, 1.5)
  )

analysis_result %>% 
  pivot_wider(
    names_from = "time",
    values_from = "mean"
  )
```

## Binding rows

Using the LotR data

First step: import each table 

```{r}
fellowship_ring = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "B3:D6") %>% 
  mutate(movie = "fellowship_ring")
```

```{r}
two_towers = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "F3:H6") %>% 
  mutate(movie = "two_towers")
```

```{r}
return_king = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "J3:L6") %>% 
  mutate(movie = "return_king")
```

Bind all the rows together

```{r binding_rows}
lotr_tidy = 
  bind_rows(fellowship_ring, two_towers, return_king) %>% 
  janitor::clean_names() %>% 
  relocate(movie) %>% 
  pivot_longer(
    female:male,
    names_to = "gender",
    values_to = "words"
  )
```

## Joining datasets

Import and clean the FAS datasets.

```{r}
pups_df =
  read_csv("data/FAS_pups.csv") %>% 
  janitor::clean_names() %>% 
  mutate(sex = recode(sex, `1` = "male", `2` = "female"))

litters_df =
  read_csv("data/FAS_litters.csv") %>% 
  janitor::clean_names() %>%  
  relocate(litter_number) %>% 
  separate(group, into = c("dose", "day_of_tx"), sep = 3)                         #seperate a column into multiple column
```

Next up, time to join them!

```{r}
fas_df = 
  left_join(pups_df, litters_df, by = "litter_number") %>% 
  arrange(litter_number) %>% 
  relocate(litter_number, dose, day_of_tx)
  
```

